---
title: "Navigating Mortgage Default Trends: <br>Insights from Fannie Mae Analysis"
author: Ayush Mishra
format: html
output: bookdown::html_document2
toc: false
editor: visual
tidy: true
---

```{=html}
<!This block assists in formatting the title, font size, chart caption, etc.– –>
<style type="text/css">
  .title {
    text-align: center;
}
body{ /* Normal  */
      font-size: 16px;
      margin: 2.5cm;
      top: 1.5cm;
      text-align: justify;
  }
  .caption {
    font-size: big;
    text-align: center;
    position: above;
}
</style>
```
```{css plotly-caption, echo = FALSE}
/*Include this block if you like the chart caption at the top instead of the bottom.*/
div.figure {
  display: table;
}
div.figure p {
  display: table-caption;
  caption-side: top;
}
```

```{r, include = FALSE}

# set working directory
setwd('/Users/ayush/Documents/MSBA_work/Sem_2/Data_Visualization/Mid-term_project/')

# import required libraries
library(plotly)
library(maps)
library(tidyverse)
library(GGally)
library(scales)
library(readr)

# Read data
data_2007<-read_rds('data_sample_2007Q4.rds')
data_2019<-read_rds('data_sample_2019Q4.rds')
def_rate <- read_csv('default_rate_ts.csv')

# Create credit groups and add year column
data_2007 <- data_2007 %>%
  mutate(credit_group = cut(CSCORE_B, 
                            breaks = c(-Inf,seq(from = 650, to = 800, by = 50), Inf),
                            labels = c('<650','651 - 700', '701 - 750', '751 - 800', '800<')))%>%
  mutate(year = 2007)

data_2019 <- data_2019 %>%
  mutate(credit_group = cut(CSCORE_B, 
                            breaks = c(-Inf,seq(from = 650, to = 800, by = 50), Inf),
                            labels = c('<650','651 - 700', '701 - 750', '751 - 800', '800<')))%>%
  mutate(year = 2019)

# merge both the dataframe
data <- rbind(data_2007, data_2019) 

```

------------------------------------------------------------------------

> # Executive Summary

This report examines the loan default trends of Fannie Mae following the
2008 housing market crash, focusing on the significance of credit scores
and loan purposes in predicting default probabilities. Analysis reveals
that loans extended to applicants with credit scores below 700
contributed significantly to defaults prior to the crash, while cash-out
refinances emerged as a notable correlate of defaults, suggesting
potential financial strain. However, by 2019, these trends had largely
corrected, indicating positive economic recovery post-crisis. The
findings underscore the importance of credit assessment and loan purpose
considerations in risk management for mortgage lenders.

> # Analysis & Observations

Fannie Mae has seen a significant drop in the default rate post the 2008
housing market crash with the current rate being lower than what was
before 2007. As seen in the Figure \@ref(fig:1) the loan default rate
peaked during the crisis rising to the all time high of 9.13% and has
slowly faded in the years to follow. From 2017 onward this change of
default rate has been almost steady and better than ever before.

```{r 1, echo=FALSE,message=FALSE, warning=FALSE, fig.height=4, fig.width=10, fig.cap="Default Rate Over Time"}

# chart 1
# time series default rate

def_rate$Date <- as.Date(def_rate$Date, format = "%m/%d/%Y")

sort_ts <- def_rate[order(def_rate$Date), ]

sort_ts %>%
  ggplot(aes(x = Date, y = `Default rate`*100)) + 
  geom_line(color = "lightblue3") + 
  geom_point(color = "lightblue3") + 
  geom_text(data = sort_ts[which.max(sort_ts$`Default rate`), ], aes(label = paste0("Highest: ", 
                                                                                    round(`Default rate` * 100, 2), "%")),
            vjust = 1, hjust = -0.1, size = 4, color = "red") +
  geom_text(data = sort_ts[which.min(sort_ts$`Default rate`), ], aes(label = paste0("Lowest: ", 
                                                                                    round(`Default rate` * 100, 2), "%")),
            vjust = -1, hjust = 1, size = 4, color = "blue") +
  geom_vline(data = rbind(sort_ts[which.max(sort_ts$`Default rate`), ], 
                          sort_ts[which.min(sort_ts$`Default rate`), ]), aes(xintercept = as.numeric(Date)),
             linetype = "dotted", color = c("red", "blue"), size = 1) +
  theme_minimal() + 
  labs(x = 'Year', 
       y = 'Default rate') + 
  theme(axis.text = element_text(face = "bold", size = 12),
        axis.title = element_text(face = "bold", size = 14),
        axis.title.x = element_text(margin = margin(t = 10)),
        legend.text = element_text(face = "bold", size = 12),
        legend.title = element_text(face = "bold", size = 14),
        panel.grid.major.y = element_line(color = "grey85"))
```

Credit score plays an important role in predicting the default
probability of a loan applicant. We group our sample data using credit
score of the primary loan applicant to analyse it correlation to
defaulting on loan. In Figure \@ref(fig:2) we can see that before the
2008 market crash a lot of loans were offered to applicant with credit
score below 700 which directly led to high default.

```{r 2, echo = FALSE, fig.width=10, fig.height = 4, fig.cap="Number Of Loans By Credit Groups"}
# chart 2
# distribution of loan by credit groups
data %>%
  select(year, LOAN_ID, credit_group, DEFAULT_FLAG) %>%
  drop_na() %>%
  group_by(year, credit_group) %>%
  mutate(num_loans = n()) %>%
  ungroup() %>%
  ggplot(aes(x = credit_group, y = num_loans, fill = as.factor(year))) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  geom_text(aes(label = num_loans),
            position = position_dodge(width = 0.8), 
            vjust = -0.5, size = 3, angle = 0, hjust = 0.5) + 
  labs(y = "Number of Loans",
       x = "Credit Group",
       fill = "Year") +
  theme_minimal() +
  theme(axis.title.x = element_text(face = "bold"),
        axis.text = element_text(face = "bold", size = 10),
        legend.position = "right",
        legend.margin = margin(t = -15),
        plot.caption = element_text(face = "italic"),
        plot.title = element_text(size = 14),
        panel.grid.minor = element_blank(),
        panel.grid.major.y = element_line(color = "grey90"),
        panel.grid.major.x = element_blank(),
        axis.text.x = element_text(angle = 0, hjust = 1)) +
  scale_fill_brewer(palette = "Paired") +
  theme(legend.key.size = unit(0.6, "cm"),
        legend.title = element_text(face = "bold"))

```

On comparing the default percentage by each credit group, in Figure
\@ref(fig:3) we can see that during 2007 many loan applicant that
defaulted had a credit score below 700.

```{r 3, warning=FALSE, message=FALSE, echo = FALSE, fig.width=10, fig.height = 4, fig.cap="Percentage Of Defaulters By Credit Groups"}
# chart 3
# Perentage of defaulters within each credit group
data %>%
  select(year, LOAN_ID, credit_group, DEFAULT_FLAG) %>%
  drop_na() %>%
  group_by(year, credit_group) %>%
  summarise(default_count = sum(DEFAULT_FLAG)) %>%
  mutate(percent = default_count / sum(default_count)) %>%
  ungroup() %>%
  ggplot(aes(x = credit_group, y = percent*100, fill = as.factor(year))) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  geom_text(aes(label = scales::percent(round(percent, 4))),
            position = position_dodge(width = 0.8), 
            vjust = -0.5, size = 3, angle = 0, hjust = 0.5) +  
  labs(y = "Percentage",
       x = "Credit Group",
       fill = "Year") +
  theme_minimal() +
  theme(axis.title.x = element_text(face = "bold"), 
        axis.text = element_text(face = "bold", size = 10),
        legend.position = "right", 
        legend.margin = margin(t = -15),
        plot.caption = element_text(face = "italic"),
        plot.title = element_text(size = 14),
        panel.grid.minor = element_blank(),
        panel.grid.major.y = element_line(color = "grey90"),
        panel.grid.major.x = element_blank(),
        axis.text.x = element_text(angle = 0, hjust = 1)) +
  scale_fill_brewer(palette = "Paired") +
  theme(legend.key.size = unit(0.6, "cm"),
        legend.title = element_text(face = "bold"))

```

The original unpaid principal balance, i.e., the dollar amount of the
loan as stated on the note at the time the loan was originated was
comparably equal for all the credit groups during the 2007, as seen in
the Figure \@ref(fig:4). However, this distribution seems to have been
corrected by the year 2019 as shown in Figure \@ref(fig:5).

```{r 4, echo = FALSE, fig.width=10, fig.height = 4, fig.cap="Distribution Of Loan Amount By Credit Group: 2007"}
# chart 3.1
# Distribution of original principal for defaulters 2007
data %>%
  filter(year == 2007, DEFAULT_FLAG == 1) %>%
  drop_na(credit_group) %>%
  ggplot(aes(x = credit_group, y = log(ORIG_UPB), color = credit_group)) +
  geom_jitter(width = 0.1, height = 0, alpha = 0.5, size = 2) +
  geom_boxplot(alpha = 0.1, outlier.color = NA) + 
  stat_summary(fun = mean, geom = "point", shape = 23, size = 2, color = "black", fill = "black") +  

  labs(x = "Credit Group", 
       y = "Log of Original Principal Amount") +
  theme_minimal() +
  theme(  # Remove ticks
        legend.position = "bottom",
        axis.title.x = element_text(face = "bold"), 
        axis.text = element_text(face = "bold", size = 10), 
        plot.caption = element_text(face = "italic"),
        plot.title = element_text(size = 10),
        panel.grid.major.y = element_line(color = "grey90")) + 
  scale_color_brewer(palette = "Set2") 
```

```{r 5, echo = FALSE, fig.width=10, fig.height = 4, fig.cap="Distribution Of Loan Amount By Credit Group: 2019"}
#chart 5
# Distribution of original principal for defaulters 2019
data %>%
  filter(year == 2019, DEFAULT_FLAG == 1) %>%
  drop_na(credit_group) %>%
  ggplot(aes(x = credit_group, y = log(ORIG_UPB), color = credit_group)) +
  geom_jitter(width = 0.1, height = 0, alpha = 0.5, size = 2) +
  geom_boxplot(alpha = 0.1, outlier.color = NA) + 
  stat_summary(fun = mean, geom = "point", shape = 23, size = 2, color = "black", fill = "black") +  

  labs(x = "Credit Group", 
       y = "Log of Original Principal Amount") +
  theme_minimal() +
  theme(
        legend.position = "bottom",
        axis.title.x = element_text(face = "bold"), 
        axis.text = element_text(face = "bold", size = 10),  
        plot.caption = element_text(face = "italic"),
        plot.title = element_text(size = 10),
        panel.grid.major.y = element_line(color = "grey90")) +
  scale_color_brewer(palette = "Set2")

```

Fannie Mae offers loans primarily to support homeownership by providing
financing for purchasing homes or refinancing existing mortgages to
promote access to affordable housing options and stimulate economic
growth.

In our analysis, the purpose of loan was categorised into 'Cash-Out
Refinance', 'Purchase' and 'Refinance'. A cash-out refinance involves
refinancing an existing mortgage for an amount greater than what is
currently owed and a traditional refinance involves replacing an
existing mortgage with a new one, typically to secure better loan terms.

The Figure \@ref(fig:6) clearly shows that 52% of the loans that
defaulted were taken to cash-out refinance. This suggests a potential
correlation between defaulting on loans and the initial purpose of
obtaining cash-out refinancing, which could be attributed to factors
such as increased indebtedness or financial hardship.

```{r 6, warning=FALSE, message=FALSE, echo = FALSE, fig.width=10, fig.height = 4, fig.cap="Distribution Of Defaulters By Loan Purpose"}

#chart 6
# distribution of loan defaulter by purpose

data %>%
  select(year, LOAN_ID, DEFAULT_FLAG, PURPOSE) %>%
  drop_na(PURPOSE) %>%
  group_by(year, PURPOSE) %>%
  summarise(default_count = sum(DEFAULT_FLAG)) %>%
  mutate(percent = default_count / sum(default_count)) %>%
  ungroup() %>%
  ggplot(aes(x = PURPOSE, y = percent*100, fill = as.factor(year))) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  geom_text(aes(label = scales::percent(round(percent, 4))),
            position = position_dodge(width = 0.8), 
            vjust = -0.5, size = 3, angle = 0, hjust = 0.5) + 
  labs(y = "Percentage Of Defaulters",
       x = "Loan Purpose",
       fill = "Year") +
  theme_minimal() +
  theme(axis.title.x = element_text(face = "bold"),
        axis.text = element_text(face = "bold", size = 10),
        legend.position = "right",
        legend.margin = margin(t = -15),
        plot.caption = element_text(face = "italic"),
        plot.title = element_text(size = 14),
        panel.grid.minor = element_blank(),
        panel.grid.major.y = element_line(color = "grey90"),
        panel.grid.major.x = element_blank(),
        axis.text.x = element_text(angle = 0, hjust = 1)) +
  scale_fill_brewer(palette = "Paired") +
  theme(legend.key.size = unit(0.6, "cm"),
        legend.title = element_text(face = "bold")) +
  scale_x_discrete(labels = c("Cash-Out Refinance", "Personal", "Refinance"))

```

Defaulting on a loan occurs when the borrower fails to meet the terms of
the loan agreement, typically by missing payments or violating other
contractual obligations. First-time defaulters have not previously
experienced a default on any of their loans and may face consequences
such as damage to their credit score, potential legal action by the
lender, and difficulty obtaining credit in the future. During the 2007
market crisis almost 92% of the loans that defaulted were the first time
defaulter which may mean that they took loan beyond their capacity or
that the overall financial market was doing bad before the it crashed in
2008. This can be seen in Figure \@ref(fig:7)

```{r 7, echo = FALSE, fig.width=10, fig.height = 4, fig.cap="Distribution Of First Time Defaulters"}

# chart 4
# distribution of first time defaulters
data %>%
  filter(DEFAULT_FLAG == 1) %>%
  group_by(year) %>%
  mutate(first_defaulter = ifelse(FIRST_FLAG == 'Y', 1, 0)) %>%
  summarise(default_count = sum(first_defaulter)) %>%
  mutate(percent = default_count / sum(default_count)) %>%
  ungroup() %>%
  ggplot(aes(x = factor(year), y = percent*100, fill = factor(year), label = paste0(round(percent * 100, 1), "%"))) +
  geom_bar(stat = "identity", width = 0.7, color = "white") +
  geom_text(position = position_stack(vjust = 0.5)) +
  scale_fill_brewer(palette = "Set1") +
  labs(x = "Year", 
       y = "Percentage", 
       fill = "Year") +
  theme_minimal() +
  theme(legend.position = "bottom",
        panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank())
  
```

During the housing market crash, which reached its peak around 2008,
several states in the United States were particularly hard-hit. Some of
the most affected states included California, Florida, Nevada, and
Arizona. These states experienced significant declines in housing
prices, high rates of foreclosure, and economic downturns due to their
reliance on real estate and construction industries. The effects of the
housing market crash varied across the country, but these states were
among the most severely impacted.

Figure \@ref(fig:8) shows that these states had the highest defaulters
for Fannie Mae while Figure \@ref(fig:9) shows it current defaulters
which is significantly low compared to 2007.

```{r 8, echo = FALSE, fig.width=10, fig.height = 4, fig.cap="Defaulter Distribution By State In 2007"}

# chart 8
# defaulters by state 2007

us_map <- data$OCC_STAT

state_defaulters <- data %>%
  filter(DEFAULT_FLAG==1, year==2007) %>%
  group_by(STATE)%>%
  summarise(defaulters = sum(DEFAULT_FLAG))

map_plot <- plot_ly(
  data = state_defaulters,
  type = "choropleth",
  locations = ~STATE,
  locationmode = "USA-states",
  z = ~defaulters,
  colorscale = "Blues2",
  text = ~paste("State: ", STATE, "<br>Defaulters: ", defaulters),
  hoverinfo = "text"
) %>%
  layout(
    geo = list(
      scope = "usa",
      projection = list(type = "albers usa"),
      showlakes = TRUE,
      lakecolor = toRGB("white")
    )
  )

# Display the interactive map
map_plot
```

```{r 9, echo = FALSE, fig.width=10, fig.height = 4, fig.cap="Defaulter Distribution By State In 2019"}

# chart 9
# defaulters by state 2019

us_map <- data$OCC_STAT

state_defaulters <- data %>%
  filter(DEFAULT_FLAG==1, year==2019) %>%
  group_by(STATE)%>%
  summarise(defaulters = sum(DEFAULT_FLAG))

map_plot <- plot_ly(
  data = state_defaulters,
  type = "choropleth",
  locations = ~STATE,
  locationmode = "USA-states",
  z = ~defaulters,
  colorscale = "Blues2",
  text = ~paste("State: ", STATE, "<br>Defaulters: ", defaulters),
  hoverinfo = "text"
) %>%
  layout(
    geo = list(
      scope = "usa",
      projection = list(type = "albers usa"),
      showlakes = TRUE,
      lakecolor = toRGB("white")
    )
  )

map_plot
```

> # Conclusion

In conclusion, the analysis of Fannie Mae's loan default rates post the
2008 housing market crash underscores the importance of credit scores
and loan purposes in predicting default probabilities. Prior to the
crash, loans extended to applicants with credit scores below 700
contributed significantly to defaults. However, by 2019, this trend had
corrected. Furthermore, the purpose of loans, particularly cash-out
refinances, showed a notable correlation with defaults, indicating
potential financial strain. Despite the severity of the housing market
crash, current default rates are considerably lower, reflecting positive
economic recovery.

(Word count: 757 )
